---
title: "HMSC_final"
output: html_notebook
---
The purpose of this R Script is to generate species detections models. We will generate multi-species habitat use models incorporating environmental covariates and species interactions. 

This script follows the technique developed by Christopher Beirne and the Wildlife Coexistence Lab at the University of British Columbia, and the WildCAM network. I will be following the stepwise process outlined in Chapter 10 of this manuscript: https://wildcolab.github.io/Introduction-to-Camera-Trap-Data-Management-and-Analysis-in-R/habitat-use.html.

Clear the working environment
```{r}
rm(list=ls())
```

Set the working directory
```{r}
setwd("/Users/kateandy/Documents/R/Riparian_Corridors")
```

Load the packages
```{r}
# Check you have them and load them
list.of.packages <- c("kableExtra", "tidyr", "ggplot2", "gridExtra", "lme4", "dplyr", "Hmsc", "jtools", "lubridate", "corrplot", "MuMIn")

new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)
```

#Set up the data frames for the model

Load the data frames
```{r}

#species list data
sp_summary <- read.csv("data/processed_data/2004250_species_list.csv")

#montly observations data
monthly_obs <- read.csv("data/processed_data/_30min_independent_monthly_observations.csv", header=T)

```

Remove species that have low detections and cannot fit well with the model
```{r}
monthly_obs <- monthly_obs[,!names(monthly_obs) %in% c("Red.Fox", "Red.Squirrel", "Puma", "Martes.Species", "Western.Spotted.Skunk", "Long.tailed.Weasel", "Stoat", "Elk", "Carolina.Flying.Squirrel", "American.Beaver", "Striped.Skunk", "Neotamias.Species", "Nutria")]


```


Remove spaces from sp_summary common names
```{r}
sp_summary$common_name<- gsub(" ", ".", sp_summary$common_name)

sp_summary <- sp_summary [(!(sp_summary$common_name == "Red.Fox") & !(sp_summary$common_name == "Red.Squirrel") & !(sp_summary$common_name == "Puma") & !(sp_summary$common_name == "Martes.Species") & !(sp_summary$common_name == "Western.Spotted.Skunk") & !(sp_summary$common_name == "Long-tailed.Weasel") & !(sp_summary$common_name == "Stoat") & !(sp_summary$common_name == "Elk") & !(sp_summary$common_name == "Carolina.Flying.Squirrel") & !(sp_summary$common_name == "American.Beaver") & !(sp_summary$common_name == "Striped.Skunk") & !(sp_summary$common_name == "Neotamias.Species") & !(sp_summary$common_name == "Nutria")),]


```

Load the camera locations covariates
```{r}
#Load the covariate dataframe
locs <- read.csv("data/processed_data/2004250_camera_locations_and_covariates.csv", header=T)

# Convert to categorical factors
locs <- locs %>% 
            mutate_if(is.character,as.factor)


#Standardize  covariates - it helps models coverage and facillitates comparison of effects sizes

library(MuMIn)
z_locs <- stdize(locs[,-(5:6)])

#join the fish columns back on to z_locs
z_locs<- left_join(z_locs,
            locs %>% dplyr::select(placename, observed_salmon_presence, recorded_salmon_presence),
            by = "placename")
```


#Build the model
```{r}
library(Hmsc)
```


Add season column to aggregate by season later 

```{r}

monthly_obs$date<- ym(monthly_obs$date)

monthly_obs$season <- "summer"
monthly_obs$season[month(monthly_obs$date) %in% c(1,2,3)] <- "winter"
monthly_obs$season[month(monthly_obs$date) %in% c(10,11,12)] <- "fall"
monthly_obs$season[month(monthly_obs$date) %in% c(4,5,6)] <- "spring"


monthly_obs<- monthly_obs[,c(1,3:17)] #change this to match the number of columns in monthly_obs

monthly_obs[is.na(monthly_obs)]<-0

library(dplyr)
monthly_obs<-monthly_obs %>%
  group_by(placename, season) %>%
  summarise_each(funs(sum))

```


Prepare the species detections dataframe. This will be our response variable dataframe in the model.
```{r}
#Subset monthly_obs to just species to make matrix
monthly_obs_sub<-monthly_obs[,c(4:16)] #change the number of columns to match subset column length

#Put the columns in alphabetical order
new_order = sort(colnames(monthly_obs_sub))
monthly_obs_sub <- monthly_obs_sub[, new_order]

# Pull the count data into its own matrix
Y <- as.matrix(monthly_obs_sub)

# Give the row names a useful label, in this case the site_date values 
# (just in case you want to check things)
#row.names(Y) <- paste(monthly_obs$placename, monthly_obs$date, sep="_")
row.names(Y) <- unique(paste(monthly_obs$placename, monthly_obs$season, sep="_"))

#Convert na's to 0 in dataframe
Y[is.na(Y)]<-0


```


Prepare the camera trap covariate dataframe.
```{r}

#join the covariate data to the observation data
Xdat <- left_join(monthly_obs[c("placename", "season", "days")], z_locs)

# All XData must be numeric or factors, so lets check what we have
#str(Xdat)
#Replace NA with 0 in any numeric column
Xdat<- Xdat %>% mutate(
    across(where(is.numeric), ~replace_na(.x, 0))
  )

```



Transform the species covariate dataframe to be used in the model.
```{r}
#Pick out the columns we want from sp_summary
species<- sp_summary[,c(6,8:22)]

#Make a log-transformed body mass column
species$log_body<- log(species$mass_g)

```

Standardize the species dataframe reformat. 
```{r}
#standardize
z_species <- stdize(species)


#order the common name column alphabetically
z_species<- z_species[order(z_species$common_name), ]

#make sure that all spaces are changed to "." instead
z_species$common_name <- gsub("'", ".", z_species$common_name)
z_species$common_name <- gsub("-", ".", z_species$common_name)

#Change rownames to species
row.names(z_species) <- paste(z_species$common_name)

#Remove common_name column
z_species<-z_species[,2:17]

#Convert na's to 0 in dataframe
z_species[is.na(z_species)]<-0

```



Set up the Bayesian framework
```{r}
nChains   = 3       # How many total repeats to run
thin      = 300       # How often to thin the samples 
samples   = 1000     # How many samples to take   
transient = 25000 # How long should the "warm up" be
verbose   = T       # Give reports on model progress


```



Set up random effects
```{r}
#Make a Site column
library(stringr) #we will need to load this package first
Xdat$Site = substr(Xdat$placename,3,6) #extracts the site name from placename
#Add a column for which city it belongs to. M = Maple Ridge. S = Squamish.
Xdat$City <- substr(Xdat$placename,1,1)

# make all character variables factors
Xdat <- Xdat %>% 
            mutate_if(is.character,as.factor)


# Add a site-level and city-level random effects (for the co-variances)
studyDesign = data.frame(station = as.factor(Xdat$Site), city = as.factor(Xdat$City))


#Assign the levels of the random effects
rL1 = HmscRandomLevel(units = studyDesign$Site)
rL2 = HmscRandomLevel(units = studyDesign$City)

```

Specify the model - mixed variables
The stream channel morphology model- remove "z.road_dist_m" from the mixed variables model
The development intensity model- remove "z.height_streambank_summer" from the mixed variables model
```{r}
# Model specification
mod <- Hmsc(Y = Y, 
            XData = Xdat[,c( "z.avg_ndvi", "z.elevation", "z.road_dist_m", "z.height_streambank_summer", "recorded_salmon_presence", "season", "habitat_type", "days", "z.average_viewshed")], 
            XFormula = ~ z.avg_ndvi + z.elevation +  z.road_dist_m + z.height_streambank_summer + recorded_salmon_presence + season + habitat_type + days + z.average_viewshed,
            TrData = z_species,
            TrFormula = ~z.log_body + z.diet_vfish + z.diet_seed + z.diet_plant0,
            studyDesign = studyDesign, 
            ranLevels = list(station = rL1, city = rL2),
            distr="poisson")


```


Fit the model
```{r}
out <-  sampleMcmc(mod, thin = thin, samples = samples, transient = transient,
               nChains = nChains, verbose = verbose)

save(out,file="rmv_sp_traits_env.Rdata")

```

----------------------- follow these steps after you have fit the model ----------------------------


#Model fit examination
```{r}
load("rmv_sp_traits_env.Rdata")
preds<- computePredictedValues(out,expected = FALSE)#computePredictedValues(out, expected = FALSE) to get Tjur's R2 and AUC
MF = evaluateModelFit(hM = out, predY = preds)
MF
```
Find the mean RMSE and SR2 values
```{r}
mean(MF$RMSE)
mean(MF$SR2)
```

```{r}
SR2<- as.data.frame(MF)
SR2$Taxonomic.Group<-unique(paste(out$spNames))


```


Stream channel morphology model results
```{r}
load("sp_traits_env_stream.Rdata")
preds<- computePredictedValues(out,expected = FALSE)#computePredictedValues(out, expected = FALSE) to get Tjur's R2 and AUC
MF = evaluateModelFit(hM = out, predY = preds)
MF
```

Mean RMSE and SR2
```{r}
mean(MF$RMSE)
mean(MF$SR2)
```

Development Intensity model results
```{r}
load("sp_traits_env_human.Rdata")
preds<- computePredictedValues(out,expected = FALSE)#computePredictedValues(out, expected = FALSE) to get Tjur's R2 and AUC
MF = evaluateModelFit(hM = out, predY = preds)
MF
```

Mean RMSE and SR2
```{r}
mean(MF$RMSE)
mean(MF$SR2)
```




Plot the SR2 to see how well the model explained for each species
```{r}

#put the values in descending order
SR2 <- SR2[order(-SR2$SR2), ]

#Now plot the results in a vertical bar chart
library(ggplot2)

ggplot(SR2, aes(x = reorder(Taxonomic.Group, SR2), y = SR2)) +
  geom_bar(stat = "identity", fill = "#808000") +
  coord_flip() +  # Flip the coordinates to create a vertical bar chart
  labs(x = "Taxonomic Group", y = "Pseudo-R2") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"))


```


Plot the variance
```{r}
head(out$X)
```

Variance figure
```{r}
out$spNames <- gsub('\\.', ' ', out$spNames)
out$spNames <- gsub("Douglas s Squirrel", "Douglas's Squirrel", out$spNames)
out$spNames <- gsub("Western Gray Squirrel", "Gray Squirrels", out$spNames)
out$rLNames<- gsub("station", "Site", out$rLNames)
out$rLNames<- gsub("city", "Study Location", out$rLNames)


plotVariancePartitioning <- function(hM, VP, cols = NULL, main = "Variance Partitioning", ...) {
   ng <- dim(VP$vals)[1]
   if (is.null(cols)) {
      cols <- heat.colors(ng, alpha = 1)
   }
   leg <- VP$groupnames
   for (r in seq_len(hM$nr)) {
      leg <- c(leg, paste("Random: ", hM$rLNames[r], sep = ""))
   }
   means <- round(100 * rowMeans(VP$vals), 1)
   for (i in 1:ng) {
      leg[i] <- paste(leg[i], " (mean = ", toString(means[i]), ")", sep = "")
   }

   # Create the barplot with thinner bars
   barplot(VP$vals, xlab = "", ylab = "Variance proportion", las = 2, cex.axis = 1, cex.names = 0.5,
          col = cols, width = 0.6, names.arg = hM$spNames, xlim = c(0,15))
   legend_x <- "right"  # Position the legend to the right
legend_y <- NULL  # Adjust the y-position as needed
legend("topright", legend = leg, fill = cols, x.intersp = 0.8, cex = 0.6, xpd = FALSE)


 mtext("Taxonomic Group", side = 1, line = 6, cex = 1, at = 4)

  
}




VP = computeVariancePartitioning(out, group = c(1,2,3,4,5,6,6,6,7,7,8,8), groupnames = c("Maximum NDVI","Elevation", "Development Intensity", "Channel Morphology", "Salmon Presence","Camera", "Habitat Type", "Season"))
par(mar = c(6, 3, 2, 5)+1)
plotVariancePartitioning(out, VP = VP, cols = c("#800000", "#e6194B", "#fabed4", "#ffd8b1", "#f58231", "#9A6324", "#808000", "#ffe119", "#669999", "#336666"))


```


Create 95% credible intervals for variable selection within model. These should not cross 0. 
```{r}
mpost2 = convertToCodaObject(out)
interval<-round(summary(mpost2$Beta[[1]], quantiles = c(0.05, 0.5, 0.95))[[2]],2)
write.csv(interval, file= "interval_final.csv")
```


#Plot the results

Create a summary plot which shows the effects of covariates on mammal detections
```{r}
#Fix up the names of the covariates
out$covNames <- gsub('z.', '', out$covNames)
out$covNames <- gsub("elevation", "Elevation", out$covNames)
out$covNames <- gsub("avg_ndvi", "Maximum NDVI", out$covNames)
out$covNames <- gsub("road_dist_m", "Road Distance", out$covNames)
out$covNames <- gsub("height_streambank_summer", "Summer Streambank Height", out$covNames)
out$covNames <- gsub("recorded_salmon_presence", "Salmon Presence", out$covNames)
out$covNames <- gsub("seasonspring", "Spring", out$covNames)
out$covNames <- gsub("seasonsummer", "Summer", out$covNames)
out$covNames <- gsub("seasonwinter", "Winter", out$covNames)
out$covNames <- gsub("habitat_typeriparian", "Riparian Habitat", out$covNames)
out$covNames <- gsub("habitat_typestream", "Stream Habitat", out$covNames)
out$covNames <- gsub("days", "Sampling effort (days)", out$covNames)
out$covNames <- gsub("average_viewshed", "Average Viewshed", out$covNames)

```



Plot species and environmental covariates
```{r}
postBeta = getPostEstimate(out, parName = "Beta")
par(mar=c(6,9,2,2))
plotBeta(out, post = postBeta, param = "Sign", supportLevel = 0.95, cex = c(0.7, 
        0.7, 0.8), mgp=c(0,1,0),  spNamesNumbers = c(T,F), covNamesNumbers = c(T, F), colors = colorRampPalette(c("#800000", "white", "#e6194B")))



```


Plot the species traits now
```{r}
plotGamma=function(hM, post, param = "Support", trOrder="Original",
  trVector= NULL, covOrder="Original", covVector=NULL, trNamesNumbers=c(TRUE,FALSE),
  covNamesNumbers=c(TRUE,FALSE), supportLevel=.9, main = NULL, cex=c(.6,.6,.8),
  colors=colorRampPalette(c("blue","white","red")), colorLevels = NULL,
  mar=c(8,9,2,0),
  smallplot=NULL, bigplot=NULL, newplot=TRUE){

   if(is.null(colorLevels)){
      if(param=="Sign"){
         colorLevels=3} else {
            colorLevels=200
         }
   }
  switch(class(hM$X)[1L],
         matrix = {
           ncolsX = ncol(hM$X)
         },
         list = {
           ncolsX = ncol(hM$X[[1]])
         }
  )

   covNames = character(hM$nc)
   for (i in 1:hM$nc) {
      sep = ""
      if (covNamesNumbers[1]) {
         covNames[i] = paste(covNames[i], hM$covNames[i], sep = sep)
         sep = " "
      }
      if (covNamesNumbers[2]) {
         covNames[i] = paste(covNames[i], sprintf("(C%d)",
            i), sep = sep)
      }
   }
   trNames = character(hM$nt)
   for (i in 1:hM$nt) {
      sep = ""
      if (trNamesNumbers[1]) {
         trNames[i] = paste(trNames[i], hM$trNames[i], sep = sep)
         sep = " "
      }
      if (trNamesNumbers[2]) {
         trNames[i] = paste(trNames[i], sprintf("(T%d)", i),
            sep = sep)
      }
   }


   if(covOrder=="Vector"){covorder=covVector}
   if(covOrder=="Original"){covorder=1:ncolsX}

   if(trOrder=="Vector"){trorder=trVector}
   if(trOrder=="Original"){trorder=1:ncol(hM$Tr)}


   mgamma=post$mean
   gammaP=post$support

   if(param=="Sign"){
      toPlot = sign(mgamma)
      toPlot = toPlot * ((gammaP>supportLevel) + (gammaP<(1-supportLevel))>0)
      gammaMat = matrix(toPlot, nrow=ncolsX, ncol=ncol(hM$Tr))
   }
   if(param=="Mean"){
      toPlot = mgamma
      toPlot = toPlot * ((gammaP>supportLevel) + (gammaP<(1-supportLevel))>0)
      gammaMat = matrix(toPlot, nrow=ncolsX, ncol=ncol(hM$Tr))
   }
   else{
      if(param=="Support"){
         toPlot = 2*gammaP-1
         toPlot = toPlot * ((gammaP>supportLevel) + (gammaP<(1-supportLevel))>0)
         gammaMat = matrix(toPlot, nrow=ncolsX, ncol=ncol(hM$Tr))
      }}

   rownames(gammaMat) = covNames
   colnames(gammaMat) = trNames
   X = gammaMat[covorder,trorder]

   old.par = par(no.readonly = TRUE)
   colors = colors(colorLevels)

   START=0
   END=.65
   ADJy=1/(ncol(X)*2)
   ADJx=1/(nrow(X)*4)

   if(newplot){
      par(fig = c(0,1,0,1),  mar = mar)
   } else {
      par(old.par, mar=mar)
   }

   plot.new()
    axis(1, at = seq(START+ADJx, END-ADJx,
                    by = ((END-ADJx) - (START+ADJx))/(nrow(X) - 1)),
         labels = FALSE)
   axis(2, at = seq(ADJy, 1-ADJy, length.out=ncol(X)), labels = FALSE)

   text(x = seq(START+ADJx, END-ADJx, by = ((END-ADJx) - (START+ADJx))/(nrow(X) - 1)), par("usr")[3] - 0.05, srt = 90, adj = 1,cex=cex[2],
      labels = covNames[covorder], xpd = TRUE)
   text(y = seq(ADJy, 1-ADJy, length.out=ncol(X)), par("usr")[3] - 0.05, srt = 0, adj = 1,cex = cex[1],
      labels = trNames[trorder], xpd = TRUE)

   if(all(is.na(X)) || sum(abs(X))==0){
      warning("nothing to plot at this level of posterior support")
      zlim = c(-1,1)
   } else{
      zlim = c(-max(abs(range(X))),max(abs(range(X))))
   }


   image.plot(x = seq(START+ADJx, END-ADJx, by = ((END-ADJx) - (START+ADJx))/(nrow(X) - 1)),
      y = seq(ADJy, 1-ADJy, length.out = ncol(X)),
      z = X, add = TRUE, nlevel = colorLevels,
      legend.width = 2, legend.mar = NULL,
      legend.cex = cex,
      axis.args=if(param=="Sign")
      {list(labels=c("+","0","-"),at=c(1,0,-1),cex.axis=cex[3],mgp=c(3,2,0),hadj=1)
      } else {
         list(cex.axis=cex[3],mgp=c(3,2,0),hadj=1)
      },
      graphics.reset = TRUE, horizontal = FALSE, bigplot = bigplot, smallplot = smallplot,
      legend.only = FALSE, col = colors,
      lab.breaks = NULL, zlim = zlim)

   if (!is.null(main))
      title(main = main)

   if(newplot){
      par(old.par)
   }
}
```


Plot the species traits and environmental predictor variables
```{r}
out$trNames <- gsub('z.', '', out$trNames)
out$trNames <- gsub("log_body", "Log-transformed Body Mass", out$trNames)
out$trNames <- gsub("diet_vfish", "Fish Diet", out$trNames)
out$trNames <- gsub("diet_seed", "Seed Diet", out$trNames)
out$trNames <- gsub("diet_plant0", "Plant Diet", out$trNames)


postGamma = getPostEstimate(out, parName = "Gamma")

par(mar=c(7,11,2.5,0))
plotGamma(out, post = postGamma, param = "Sign",  supportLevel = 0.95,  colors = colorRampPalette(c("#800000", "white", "#e6194B")))

```

Create a plot which shows the interaction effect between species.
```{r}
OmegaCor = computeAssociations(out)

supportLevel = 0.0

toPlot = ((OmegaCor[[2]]$support>supportLevel)
+ (OmegaCor[[2]]$support<(1-supportLevel))>0)*OmegaCor[[2]]$mean #change all the 1s to 2s for city random effect


corrplot(toPlot, 
         method = "color",
         type="upper",
         order = "FPC",
        colors = colorRampPalette(c("#800000", "white", "#e6194B"))(200),
        #title = paste("Random Effect Level:", 
                      #mod$rLNames[1]), mar=c(0,0,1,0), #change to mod$rLNames[1] for station and mod$rlNames[2] for city
        tl.col="black",
        tl.cex = 0.5,
         number.font = 0.5,
         cl.ratio = .2)

#median(toPlot)
#sd(toPlot)
```

















