---
title: "Principal Component Analysis"
output: html_notebook
author: Kate Andy
created: June 6, 2023
---

The purpose of this R Notebook is to run principal component analyses/principal coordinate analyses of various vegetation and habitat characteristic data collected throughout the duration of the Riparian Corridors project. These data will be incorporated into the main covariates dataframe.  

Clear the workspace and set the working directory
```{r}
rm(list=ls())
setwd("/Users/kateandy/Documents/R/Riparian_Corridors")
```


Load the required packages
```{r}
library(readr)
library(vegan)
library(dplyr)
library(ggfortify)
library(ggplot2)
library(ecodist)
```


Read in datasets - these data come from seasonal field work. 
```{r}
quadratCover <- read_csv("PCA_data/quadrat_cover.csv")
streamQuadratCover<- read_csv("PCA_data/stream_quadrat_cover.csv")
terrestrialSpecies<- read_csv("PCA_data/total_terrestrial_species_composition.csv")
treeBiomass<- read_csv("PCA_data/tree_biomass.csv")
treeDensity<- read_csv("PCA_data/tree_density.csv")
```



Run the PCA for the quadrat cover dataset
```{r}
#Pull out just summer quadrat data - choosing maximum growing period
quadratSummer<-filter(quadratCover, season_visit == "Jun-22")

#Summarize so that there is only average quadrat values per site.
quadratSummergroup<- quadratSummer %>% group_by(site, habitat_type, location) %>% summarise(across(where(is.numeric), list(mean=mean), na.rm=TRUE))

#Remove the replicate number column
quadratSummergroup<-quadratSummergroup[,-4]
  
#Subset data so it is just values for PCA
quadrat_data <- quadratSummergroup[, 4:16]

#Scale and center all the data
quadrat_scaled<-scale(quadrat_data, center = TRUE, scale = TRUE)

#Run the PCA
quadrat_PCA <- princomp(quadrat_scaled, cor = FALSE)

```

Plot the results from the PCA.
```{r}
autoplot(quadrat_PCA, data = quadratSummergroup, colour = 'habitat_type', shape = 'location', loadings=TRUE,loadings.label = TRUE, loadings.label.size = 3)
```


Interpret the results
```{r}
summary(quadrat_PCA)
loadings(quadrat_PCA)
```

Extract quadrat scores for Comp 1 and Comp 2
```{r}
#Create a new vector called "Sites"
Sites<-quadratSummergroup$site 

#Create a vector of the PCA Comp1 scores
Comp1<-quadrat_PCA$scores[,"Comp.1"]

#Create a vector of the PCA Comp2 scores
Comp2<- quadrat_PCA$scores[,"Comp.2"]

#Merge the vectors to create a dataframe
quadratVars<-data.frame(Sites,Comp1,Comp2)
```


Run the PCA for the stream quadrat cover dataset
```{r}
#Pull out just summer quadrat data - choosing maximum growing period
streamQuadratSummer<-filter(streamQuadratCover, season_visit == "22-Jun") #sometimes needs to be "Jun-22" instead of "22-Jun"

#Summarize so that there is only average quadrat values per site.
streamQuadratSummergroup<- streamQuadratSummer %>% group_by(site, location) %>% summarise(across(where(is.numeric), list(mean=mean), na.rm=TRUE))

#Remove the replicate number column
streamQuadratSummergroup<-streamQuadratSummergroup[,-3]

#Subset data so it is just values for PCA
stream_quadrat_data <- streamQuadratSummergroup[, 3:12]

#Scale and center all the data
stream_quadrat_scaled<-scale(stream_quadrat_data, center = TRUE, scale = TRUE)

#Replace NA values with 0
stream_quadrat_scaled[is.na(stream_quadrat_scaled)] <- 0

#Run the PCA
stream_quadrat_PCA <- princomp(stream_quadrat_scaled, cor = FALSE)
```

Plot the results from the PCA
```{r}
autoplot(stream_quadrat_PCA, data = streamQuadratSummergroup, colour = 'location') + geom_text(size = 2, vjust = -1, label = streamQuadratSummergroup$site)
```
Interpret the results
```{r}
summary(stream_quadrat_PCA)
loadings(stream_quadrat_PCA)
```

Extract stream quadrat scores for Comp 1 and Comp 2
```{r}
#Create a new vector of the camera trap sites
Sites<-streamQuadratSummergroup$site

#Create a vector of the PCA Comp1 scores
Comp1<-stream_quadrat_PCA$scores[,"Comp.1"]

#Create a vector of the PCA Comp2 scores
Comp2<- stream_quadrat_PCA$scores[,"Comp.2"]

#Merge the vectors to make a dataframe
streamQuadratVars<-data.frame(Sites,Comp1,Comp2)
```


Run the PCoA for the terrestrial species dataset
```{r}
#Subset data so it is just values for PCoA
terrestrialSpecies_data <- terrestrialSpecies[, 4:54]

#Run the PCoA
terrestrialSpecies.bray <- vegdist(terrestrialSpecies_data, method = "bray") #creates a dissimilarity matrix
terrestrial_species_PCoA <- pco(terrestrialSpecies.bray, negvals = "zero", dround = 0) #runs the PCoA

```

Plot the results from the PCoA
```{r}
#Extract PCoA1 and PCoA2 vectors
PCoA1<- terrestrial_species_PCoA$vectors[,1]
PCoA2<-terrestrial_species_PCoA$vectors[,2]

#Merge into a dataframe to plot
PCoA<-data.frame(PCoA1,PCoA2)

#Create the plot
ggplot(PCoA,aes(x= PCoA1, y= PCoA2, color = terrestrialSpecies$habitat_type, shape = terrestrialSpecies$location)) +
  geom_point()+ 
  geom_segment(data = loadings,
                 aes(xend = PCoA1, yend = PCoA2, x = 0, y = 0),
                 arrow = arrow(type = "closed", length = unit(0.1, "inches"), ends = "both"))



```



Interpret the results
```{r}
summary(terrestrial_species_PCoA)
loadings(terrestrial_species_PCoA)
```

Extract terrestrial species vectors for PCoA 1 and PCoA 2
```{r}
#Create a vector of camera trap sites
Sites<-terrestrialSpecies$site

#Create a vector of the PCoA1 vectors
PCoA1<- terrestrial_species_PCoA$vectors[,1]

#Create a vector of the PCoA2 vectors
PCoA2<-terrestrial_species_PCoA$vectors[,2]

#Merge the vectors into a dataframe
terrestrialVars<-data.frame(Sites,PCoA1,PCoA2)
```


Run the PCA for the tree density dataset
```{r}
#Subset data so it is just values for PCA
treeDensity_data <- treeDensity[, 4:20]

#Scale and center all the data
tree_density_scaled<-scale(treeDensity_data, center = TRUE, scale = TRUE)

#Run the PCA
tree_density_PCA <- princomp(tree_density_scaled, cor = FALSE)
```

Plot the results from the PCA
```{r}
autoplot(tree_density_PCA, data = treeDensity, colour = 'habitat_type', shape = 'location')
```
Interpret the results
```{r}
summary(tree_density_PCA)
loadings(tree_density_PCA)
```

Extract density scores for Comp 1 and Comp 2
```{r}
Sites<-treeDensity$site
Comp1<-tree_density_PCA$scores[,"Comp.1"]
Comp2<- tree_density_PCA$scores[,"Comp.2"]
densityVars<-data.frame(Sites,Comp1,Comp2)
```



Run the PCA for the tree biomass dataset
```{r}
#Subset data so it is just values for PCA
treeBiomass_data <- treeBiomass[, 4:19]
#Replace NA values with 0
treeBiomass_data[is.na(treeBiomass_data)] <- 0

#Scale and center all the data
tree_biomass_scaled<-scale(treeBiomass_data, center = TRUE, scale = TRUE)

#Run the PCA
tree_biomass_PCA <- princomp(tree_biomass_scaled, cor = FALSE)
```

Plot the results from the PCA
```{r}
autoplot(tree_biomass_PCA, data = treeBiomass, colour = 'habitat_type', shape = 'location')
```
Interpret the results
```{r}
summary(tree_biomass_PCA)
loadings(tree_biomass_PCA)
```

Extract biomass scores for Comp 1 and Comp 2
```{r}
#Create a vector of the camera trap sites
Sites<-treeBiomass$site

#Create a vector of the PCA Comp1 scores
Comp1<-tree_biomass_PCA$scores[,"Comp.1"]

#Create a vector of the PCA Comp2 scores
Comp2<- tree_biomass_PCA$scores[,"Comp.2"]

#Merge the vectors to make a dataframe
biomassVars<-data.frame(Sites,Comp1,Comp2)
```


Export PCA/PCoA variables
```{r}
setwd("~/Documents/R/Riparian_Corridors/PCA_data")
write.csv(biomassVars, file = "biomassVars.csv")
write.csv(quadratVars, file = "quadratVars.csv")
write.csv(streamQuadratVars, file = "streamQuadratVars.csv")
write.csv(terrestrialVars, file = "terrestrialVars.csv")
```

These data will be incorporated into the main covariates dataframe.


----------------------END------------------------------