---
title: "Summary of Vegetation Survey Data"
output: html_notebook
author: Kate Andy
date of last edit: March 2, 2023
---
The purpose of this R Notebook is to summarize the data collected from vegetation surveys over the season so I can add them to the master covariates data table.

Clear the workspace
```{r}
rm(list=ls())
```

Set the working directory
```{r}
setwd("/Users/kateandy/Documents/R/Riparian_Corridors")
```

Load the necessary packages
```{r}
library(dplyr)
library(readr)
```

Load the seaonsal data for Squamish and Maple Ridge
```{r}
#Stream Summer Maple Ridge
StreamSummerMR<- read_csv("Veg_Surveys/Veg_survey_summer_MR_stream.csv")

#Non-stream Summer Maple Ridge
nonStreamSummerMR<- read_csv("Veg_Surveys/Veg_survey_summer_MR_nonStream.csv")

#Stream Fall Maple Ridge
StreamFallMR <- read_csv("Veg_Surveys/Veg_survey_fall_MR_stream.csv")

#Non-stream Fall Maple Ridge
nonStreamFallMR <- read_csv("Veg_Surveys/Veg_survey_fall_MR_nonStream.csv")

#Stream Summer Squamish
StreamSummerSquam <- read_csv("Veg_Surveys/Veg_survey_summer_Squamish_stream.csv")

#Non-stream Summer Squamish
nonStreamSummerSquam <- read_csv("Veg_Surveys/Veg_survey_summer_Squamish_nonStream.csv")

#Stream Fall Squamish
StreamFallSquam <- read_csv("Veg_Surveys/Veg_survey_fall_Squamish_stream.csv")

#Non-stream Fall Squamish
nonStreamFallSquam<- read_csv("Veg_Surveys/Veg_survey_fall_Squamish_nonStream.csv")

#Stream Winter Squamish
StreamWinterSquam <-  read_csv("Veg_Surveys/Veg_survey_winter_Squamish_stream.csv")

#Non-stream Winter Squamish
nonStreamWinterSquam <-read_csv("Veg_Surveys/Veg_survey_winter_Squamish_nonStream.csv")

#Stream Winter Maple Ridge
StreamWinterMR<- read_csv("Veg_Surveys/Veg_survey_winter_MR_stream.csv")

#NOn-stream Winter Maple Ridge
nonStreamWinterMR <- read_csv("Veg_Surveys/Veg_survey_winter_MR_nonStream.csv")

#Stream Spring Squamish
StreamSpringSquam<- read_csv("Veg_Surveys/Veg_survey_spring_Squamish_stream.csv")

#Non-stream Spring Squamish
nonStreamSpringSquam<- read_csv("Veg_Surveys/Veg_survey_spring_Squamish_nonStream.csv")

#Stream Spring Maple Ridge
StreamSpringMR<- read_csv("Veg_Surveys/Veg_survey_spring_MR_stream.csv")

#Non-stream Spring Maple Ridge
nonStreamSpringMR<- read_csv("Veg_Surveys/Veg_survey_spring_MR_nonStream.csv")
```


Summarize the data for the average depth of duff per site
```{r}
#Summer Squamish
DuffSummerSquam <- nonStreamSummerSquam %>% group_by(`Site Name`) %>% summarise(
  Duff_avg = mean(`Depth of Duff`)
)

#Summer Maple Ridge
DuffSummerMR <- nonStreamSummerMR %>% group_by(`Site Name`) %>% summarise(
  Duff_avg = mean(`Depth of Duff`)
)

#Fall Squamish
DuffFallSquam<- nonStreamFallSquam %>% group_by(`Site Name`) %>% summarise(
  Duff_avg = mean(`Depth of Duff`)
)

#Fall Maple Ridge
DuffFallMR <- nonStreamFallMR %>% group_by(`Site Name`) %>% summarise(
  Duff_avg = mean(`Depth of Duff`)
)

#Winter Squamish
DuffWinterSquam<- nonStreamWinterSquam %>% group_by(`Site Name`) %>% summarise(
  Duff_avg = mean(`Depth of Duff`)
)

#Winter Maple Ridge
DuffWinterMR<- nonStreamWinterMR %>% group_by(`Site Name`) %>% summarise(
  Duff_avg = mean(`Depth of Duff`)
)

#Spring Squamish
DuffSpringSquam<- nonStreamSpringSquam %>% group_by(`Site Name`) %>% summarise(
  Duff_avg = mean(`Depth of Duff`)
)

#Spring Maple Ridge
DuffSpringMR<- nonStreamSpringMR %>% group_by(`Site Name`) %>% summarise(
  Duff_avg = mean(`Depth of Duff`)
)
```


Summarize the data for the average tallest woody and herbaceous height
```{r}
#Summer Squamish
#get an average of the woody values
nonStreamSummerSquam$Average_woody_height<- rowMeans(nonStreamSummerSquam[,c(12:17)])

#get an average of the herbaceous values
nonStreamSummerSquam$Average_herbaceous_height<- rowMeans(nonStreamSummerSquam[,c(18:23)])

#summarize the data by mean woody and herbaceous height
VegHeightSummerSquam<- nonStreamSummerSquam %>% group_by(`Site Name`) %>% summarise(
  Woody = mean(Average_woody_height), 
  Herbaceous = mean(Average_herbaceous_height)
)

#Summer Maple Ridge
#get an average of the woody values
nonStreamSummerMR$Average_woody_height<- rowMeans(nonStreamSummerMR[,c(12:17)])

#get an average of the herbaceous values
nonStreamSummerMR$Average_herbaceous_height<- rowMeans(nonStreamSummerMR[,c(18:23)])

#summarize the data by mean woody and herbaceous height
VegHeightSummerMR<- nonStreamSummerMR %>% group_by(`Site Name`) %>% summarise(
  Woody = mean(Average_woody_height), 
  Herbaceous = mean(Average_herbaceous_height)
)

#Fall Squamish
#get an average of the woody values
nonStreamFallSquam$Average_woody_height<- rowMeans(nonStreamFallSquam[,c(12:17)])

#get an average of the herbaceous values
nonStreamFallSquam$Average_herbaceous_height<- rowMeans(nonStreamFallSquam[,c(18:23)])

#summarize the data by mean woody and herbaceous height
VegHeightFallSquam<- nonStreamFallSquam %>% group_by(`Site Name`) %>% summarise(
  Woody = mean(Average_woody_height), 
  Herbaceous = mean(Average_herbaceous_height)
)

#Fall Maple Ridge
#get an average of the woody values
nonStreamFallMR$Average_woody_height<- rowMeans(nonStreamFallMR[,c(12:17)])

#get an average of the herbaceous values
nonStreamFallMR$Average_herbaceous_height<- rowMeans(nonStreamFallMR[,c(18:23)])

#summarize the data by mean woody and herbaceous height
VegHeightFallMR<- nonStreamFallMR %>% group_by(`Site Name`) %>% summarise(
  Woody = mean(Average_woody_height), 
  Herbaceous = mean(Average_herbaceous_height)
)

#Winter Squamish
#get an average of the woody values - calling 11-16 columns because no 'predicted line of movement' column, unlike in the fall and summer dataframe
nonStreamWinterSquam$Average_woody_height<- rowMeans(nonStreamWinterSquam[,c(11:16)])

#get an average of the herbaceous values
nonStreamWinterSquam$Average_herbaceous_height<- rowMeans(nonStreamWinterSquam[,c(17:22)])

#summarize the data by mean woody and herbaceous height
VegHeightWinterSquam<- nonStreamWinterSquam %>% group_by(`Site Name`) %>% summarise(
  Woody = mean(Average_woody_height), 
  Herbaceous = mean(Average_herbaceous_height)
)

#Winter Maple Ridge
#get an average of the woody values - calling 11-16 columns because no 'predicted line of movement' column, unlike in the fall and summer dataframe
nonStreamWinterMR$Average_woody_height<- rowMeans(nonStreamWinterMR[,c(11:16)])

#get an average of the herbaceous values
nonStreamWinterMR$Average_herbaceous_height<- rowMeans(nonStreamWinterMR[,c(17:22)])

#summarize the data by mean woody and herbaceous height
VegHeightWinterMR<- nonStreamWinterMR %>% group_by(`Site Name`) %>% summarise(
  Woody = mean(Average_woody_height), 
  Herbaceous = mean(Average_herbaceous_height)
)

#Spring Squamish
#get an average of the woody values - calling 11-16 columns because no 'predicted line of movement' column, unlike in the fall and summer dataframe
nonStreamSpringSquam$Average_woody_height<- rowMeans(nonStreamSpringSquam[,c(11:16)])

#get an average of the herbaceous values
nonStreamSpringSquam$Average_herbaceous_height<- rowMeans(nonStreamSpringSquam[,c(17:22)])

#summarize the data by mean woody and herbaceous height
VegHeightSpringSquam<- nonStreamSpringSquam %>% group_by(`Site Name`) %>% summarise(
  Woody = mean(Average_woody_height), 
  Herbaceous = mean(Average_herbaceous_height)
)

#Spring Maple Ridge
#get an average of the woody values - calling 11-16 columns because no 'predicted line of movement' column, unlike in the fall and summer dataframe
nonStreamSpringMR$Average_woody_height<- rowMeans(nonStreamSpringMR[,c(11:16)])

#get an average of the herbaceous values
nonStreamSpringMR$Average_herbaceous_height<- rowMeans(nonStreamSpringMR[,c(17:22)])

#summarize the data by mean woody and herbaceous height
VegHeightSpringMR<- nonStreamSpringMR %>% group_by(`Site Name`) %>% summarise(
  Woody = mean(Average_woody_height), 
  Herbaceous = mean(Average_herbaceous_height)
)
```


Summarize for mean canopy cover
```{r}
#Summer Maple Ridge non-stream
CanopySummerMR<- nonStreamSummerMR %>% group_by(`Site Name`) %>% summarise(
  Canopy = mean(`Canopy Cover`)
)

#Summer Maple Ridge stream
CanopySSummerMR<- StreamSummerMR %>% group_by(`Site Name`) %>% summarise(
  Canopy = mean(`Canopy Cover`)
)

#Summer Squamish non-stream
CanopySummeSquam <- nonStreamSummerSquam %>% group_by(`Site Name`) %>% summarise(
  Canopy = mean(`Canopy Cover`)
)

#Summer Squamish stream
CanopySSummerSquam<- StreamSummerSquam %>% group_by(`Site Name`) %>% summarise(
  Canopy = mean(`Canopy Cover`)
)

#Fall Maple Ridge non-stream
CanopyFallMR<- nonStreamFallMR %>% group_by(`Site Name`) %>% summarise(
  Canopy = mean(`Canopy Cover`)
)

#Fall Maple Ridge stream
CanopySFallMR <- StreamFallMR %>% group_by(`Site Name`) %>% summarise(
  Canopy = mean(`Canopy Cover`)
)

#Fall Squamish non-stream
CanopyFallSquam<- nonStreamFallSquam %>% group_by(`Site Name`) %>% summarise(
  Canopy = mean(`Canopy Cover`)
)

#Fall Squamish stream
CanopySFallSquam<- StreamFallSquam %>% group_by(`Site Name`) %>% summarise(
  Canopy = mean(`Canopy Cover`)
)

#Winter Maple Ridge non-stream
CanopyWinterMR<- nonStreamWinterMR %>% group_by(`Site Name`) %>% summarise(
  Canopy = mean(`Canopy Cover`)
)

#Winter Maple Ridge stream
CanopySWinterMR <- StreamWinterMR %>% group_by(`Site Name`) %>% summarise(
  Canopy = mean(`Canopy Cover`)
)

#Winter Squamish non-stream
CanopyWinterSquam<- nonStreamWinterSquam %>% group_by(`Site Name`) %>% summarise(
  Canopy = mean(`Canopy Cover`)
)

#Winter Squamish stream
CanopySWinterSquam<- StreamWinterSquam %>% group_by(`Site Name`) %>% summarise(
  Canopy = mean(`Canopy Cover`)
)

#Spring Squamish non-stream
CanopySpringSquam<- nonStreamSpringSquam %>% group_by(`Site Name`) %>% summarise(
  Canopy = mean(`Canopy Cover`)
)

#Spring Squamish stream
CanopySSpringSquam<- StreamSpringSquam %>% group_by(`Site Name`) %>% summarise(
  Canopy = mean(`Canopy Cover`)
)

#Spring Maple Ridge non-stream
CanopySpringMR<- nonStreamSpringMR %>% group_by(`Site Name`) %>% summarise(
  Canopy = mean(`Canopy Cover`)
)

#Spring Maple Ridge stream
CanopySSpringMR <- StreamSpringMR %>% group_by(`Site Name`) %>% summarise(
  Canopy = mean(`Canopy Cover`)
)


```


Now let's bring in the tree data and look at the total DBH and dominant species
```{r}
Trees<- read_csv("Veg_Surveys/Tree_Inventory.csv")
```

Summarize by total DBH and dominant species
```{r}
#First, remove the shrubs from our tree list
treesFinal<-Trees[!(Trees$`Tree Species`=="vine maple" | Trees$`Tree Species`=="pacific crab apple"| Trees$`Tree Species`=="common hawthorn" | Trees$`Tree Species`=="choke cherry" | Trees$`Tree Species`=="bitter cherry" | Trees$`Tree Species`=="red osier dogwood"),]

#Summarize by total DBH and dominant species
DBHSummary<- treesFinal %>% group_by(Site) %>% summarise(
  DBH_tot = sum(DBH, na.rm = TRUE)
)

#export csv
write.csv(DBHSummary, file = "DBHSummary.csv")

#Make basal area a numeric variable
treesFinal$`Basal Area per Tree`<-as.numeric(treesFinal$`Basal Area per Tree`)

#Summarize frequency of tree species, total DBH, and total basal area by site
TreeSummary<- treesFinal %>% group_by(Site, `Tree Species`) %>% summarise(
  Frequency = n(),
  DBH = sum(DBH),
  Area = sum(`Basal Area per Tree`)
) 

#now just pull out the top 3 species per site
TreeSummary<- TreeSummary %>% top_n(4,Frequency)

#Look at the unique tree species
Species<- unique(TreeSummary$`Tree Species`)
Species


```

Summarize shrub inventory data
```{r}
Shrubs <- read.csv("Veg_Surveys/Shrub_Inventory.csv")

#Look at the unique shrub species
ShrubSpecies<- unique(Shrubs$Shrub.Species)
ShrubSpecies

ShrubsGroup<-Shrubs %>% group_by(Shrub.Species)
```

To determine the stream size classification, let's have a look at the distribution of bankfull widths across all sites
```{r}
covariates <- read_csv("data/covariates.csv")
```

Subset to the bankfull observations
```{r}
#Extract the variables desired 
bankfull<-covariates[,c("placename", "bankfull_width_summer", "bankfull_width_winter", "bankfull_width_spring")]

#Find the bankfull means
bankfull$mean<-rowMeans(bankfull[,2:4], na.rm = TRUE)

#Summarize the bankfull means and range
bankfullSummary<- bankfull[,c(1,5)]
range(bankfullSummary$mean)


#plot the distribution of means
library(ggplot2)
ggplot(bankfullSummary, aes(x=mean)) + geom_histogram(color="black", fill="white", binwidth = 1)

```


------------------------ END ----------------------

